Liveness-Based Garbage Collection for Lazy Languages

### 2016-02-13 paper/paper.tex
nearly there -- 12 pages

\documentclass[9pt]{sigplanconf}
\usepackage{etex}
\usepackage{float}
\usepackage{mathrsfs}
%%\usepackage{lineno}
\usepackage{booktabs}
\usepackage{stmaryrd}
\usepackage{boxedminipage}
\usepackage[T1]{fontenc}
\usepackage{epsfig}
\usepackage{multirow}
\usepackage{url}
\usepackage[normalem]{ulem}
\usepackage{schemepgm}
\usepackage{graphicx}
\usepackage{graphpap}
\usepackage{tabularx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{pstricks}
\usepackage{pst-text}
\usepackage{pst-node}
\usepackage{pst-tree}
\usepackage{pst-rel-points}
\usepackage{bcprules}
\usepackage{subfigure}
\usepackage[boxed]{algorithm2e}
\usepackage{makecell}
\usepackage{balance}

@@

In  this work  we show  that  static analysis  of heap  data can  help
garbage  collectors   in  reclaiming   more  garbage.    The  specific
contributions of this paper are
\added{
\begin{itemize}
\item We  propose a liveness  based GC scheme  for a lazy  first order
  functional language and  prove its correctness.  To the  best of our
  knowledge,  this is  the first  work on  liveness-based GC  for lazy
  languages.       All      previous      attempts~\cite{shaham01heap,
    ran.shaham-sas03,          shaham02estimating,         asati14lgc,
    karkare06effectiveness} have  involved either imperative  or eager
  functional languages.  The novelty of our approach is in handling of
  liveness  of closures,  an  issue  that arises  due  to laziness  at
  runtime.

@@

A notion  that generalizes liveness  is {\em demand}.   While liveness
gives the patterns of future uses of a variable, demand represents the
future uses of the value of an expression.  The demand on an expression
$e$   is   again   a   set    of   access   paths---that   subset   of
$\{\acar,\acdr\}^\ast$ which the  context of $e$ may  explore of $e$'s
result.   To see  the need  for demands,  consider the expression
$(\LET~x\leftarrow  (\CDR~y)~\IN~   (\SRETURN~x))$.   Assume   that  the
context of this expression places  the demand $\lbrace \epsilon, \acar
\rbrace$. Since the  value of the expression is the  value of $x$, the
demand translates to the liveness  $[x \mapsto \lbrace \epsilon, \acar
  \rbrace]$.   Due  to  the  \LET\   definition  which  binds  $x$  to
$(\CDR~y)$, the liveness of $x$  now becomes the demand on $(\CDR~y)$.
This, in  turn, generates  the liveness $\lbrace  y.\epsilon, y.\acdr,
y.\acdr\acar  \rbrace$.   These are  the  set  of $y$-rooted  accesses
required to  explore $\lbrace  \epsilon, \acar  \rbrace$ paths  of the
result of $(\CDR~y)$.
%% As an  analogy with classical  (strong) liveness analysis,  $y$ and
%% $z$ are live at the entry $\pi: x:=y+z$, if and only if $x$ is live
%% at  exit  of $\pi$.   In  our  terminology, the  liveness  $\lbrace
%% \epsilon\rbrace $ of $x$ at the  exit from $\pi$ becomes the demand
%% on  $y+z$,  and  this,  in turn  generates  the  liveness  $\lbrace
%% y.\epsilon, z.\epsilon \rbrace$ at the entry of $\pi$.

@@

\begin{figure}[t!]
\begin{center}\footnotesize
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{|@{}c@{\ }|@{\ }c@{\ }|@{\ }c@{\ }|}
\hline
Premise & Transition & Rule name \\ 
\hline
\hline 
\makecell[t]{ $\rho(x) \mbox{ is } \bot$} & $\rho, S,
  \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow \bang$ &
{\sc car-bang} 
\\
\hline
\makecell[t]{$\heap(\rho(x)) \mbox{ is } (\langle s, \rho'\rangle, d)$} & \makecell[t]{$\rho,\, S,\,  \heap,\,
(\CAR~x), \cred{\sigma} \rightsquigarrow$ \\ $ \rho', \,(\rho, addr(\langle
s, \rho'\rangle), (\CAR~x),\cred{\sigma} )\!:\!S,$\\ $\quad \heap,\, s, \, \cred{\sigma}$ }     &
{\sc car-1-clo} 
\\
\hline
\makecell[t]{ $\heap(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$} & 
\makecell[t]{ $\rho, S, \heap, (\CAR~x), \cred{\sigma}
  \rightsquigarrow$ \\
  $\rho', (\rho, \rho(x), (\CAR~x), \cred{\sigma})\!:\!S,$\\ $\quad \heap, s,\cred{\renewcommand{\arraystretch}{1}\left\{
    \begin{array}{lr}
      {\emptyset}&\mbox{if } \sigma = \emptyset \\       
      {\{\epsilon\} \cup \acar\sigma }&\mbox{otherwise}
    \end{array}\right.
  }$   }      
&
{\sc car-clo}\\
\hline
\makecell[t]{$\cred{GC(\rho_1, S_1, \heap_1, (\LET~x\leftarrow
    s~\IN~e), \sigma)}$\\ $\cred{ = (\rho, S, \heap)}$,\\$\ell$ is a new location}& \makecell[t]{$\rho, S, \heap, (\LET~x\leftarrow
  s~\IN~e), \cred{\sigma}$  $ \rightsquigarrow$ \\ $ \rho\oplus[x
    \mapsto \ell], S,$\\$ \heap[\ell := \langle s, \lfloor\rho\rfloor_{FV(s)}, \sigma_x\rangle], e, \cred{\sigma}$ \\
    where $\sigma_x\ =\  \lfloor\mathcal{L}(e,\sigma,\Lfonly)\rfloor_{\{x\}}$} &
{\sc let} \\ 
\hline
\end{tabular}
\end{center}
\caption{Minefield  semantics for  \CAR\   and \LET.
  \warning{Add a note mentioning that we handle the issue of requiring
    a      specific     demand      by      introducing     a      new
    symbol}\label{fig:minefield-semantics-for-some}}
\end{figure}

@@

\begin{figure}[t]
  \begin{minipage}{.45\columnwidth}
    \begin{align*}
      \nont{i}{\acar} &\rightarrow \acar\nont{j}{\acar}\bcar\nont{\mathit{all}}{} \\
      \nont{i}{\acar} &\rightarrow \clazy\nont{j}{\acar}\bcar\nont{\mathit{all}}{} \\
      \nont{\mathit{all}}{} &\rightarrow \epsilon \mid \acar\nont{\mathit{all}}{} \mid
      \acdr\nont{\mathit{all}}{}
    \end{align*}
  \end{minipage}
  \begin{minipage}{.45\columnwidth}
    \input{equivalent-program}
  \end{minipage}\\
  \mbox{}\hfill(a)\hfill(b)\hfill\mbox{}
  \caption{A possible grammar generated by the proof of
    Lemma~\ref{lemma:grammar-undecidable} and a program to realize the
  grammar.}\label{fig:turing-pgm}
\vspace*{-5mm}
\end{figure}

@@

\begin{figure*}[t!]
\renewcommand{\arraystretch}{.1}
\begin{tabular}{@{}c@{}@{}c@{}@{}c@{}}
   \hskip -4mm{\epsfig{file=fibheap.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=sudoku.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=nperm.eps, height=\hgt}}
\\ \hskip -4mm{\epsfig{file=paraffins.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=lcss.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=huffman.eps, height=\hgt}}
\\ \hskip -4mm{\epsfig{file=knightstour.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=nqueens.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=deriv.eps, height=\hgt}}
\\ \hskip -4mm{\epsfig{file=treejoin.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=lambda.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=gc_bench.eps, height=\hgt}}
%\\ (a) & (b)
\end{tabular}%\vskip -10mm
 \caption{Memory usage.  
%% Column (a) shows complete  usage, column 
%%    (b) enlarges  a part  of the  usage.  
The \RGCLine\ and the \LGCLine\ curves indicate the number of cons
cells  in  the  active   semi-space  for  RGC  and  LGC
respectively.  The \ReachLine\ curve represents the number of
reachable cells and the  \UseLine\ curve represents the
number  of  cells  that  are actually  live  (of  which
liveness analysis does a static approximation).  x-axis
is the time measured  in number of cons-cells allocated
(scaled down by factor $10^5$). y-axis is the number of
cons-cells (scaled down by $10^3$).}
\label{fig:memory-usage} \figrule
\end{figure*}

\vspace*{-2.5mm}
\section{The GC scheme}
\label{sec:GC-scheme}

>>>>

\documentclass[9pt,preprint]{sigplanconf}
\usepackage{etex}
\usepackage{float}
\usepackage{mathrsfs}
%%\usepackage{lineno}
\usepackage{booktabs}
\usepackage{stmaryrd}
\usepackage{boxedminipage}
\usepackage[T1]{fontenc}
\usepackage{epsfig}
\usepackage{multirow}
\usepackage{url}
\usepackage[normalem]{ulem}
\usepackage{schemepgm}
\usepackage{graphicx}
\usepackage{graphpap}
\usepackage{tabularx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{pstricks}
\usepackage{pst-text}
\usepackage{pst-node}
\usepackage{pst-tree}
\usepackage{pst-rel-points}
\usepackage{bcprules}
\usepackage{subfigure}
\usepackage[boxed]{algorithm2e}
\usepackage{makecell}
\usepackage{balance}

@@

In  this work  we show  that  static analysis  of heap  data can  help
garbage  collectors   in  reclaiming   more  garbage.    The  specific
contributions of this paper are:
\added{
\begin{itemize}
\item We  propose a liveness  based GC scheme  for a lazy  first order
  functional language and  prove its correctness.  To the  best of our
  knowledge,  this is  the first  work on  liveness-based GC  for lazy
  languages.       All      previous      attempts~\cite{shaham01heap,
    ran.shaham-sas03,          shaham02estimating,         asati14lgc,
    karkare06effectiveness} have  involved either imperative  or eager
  functional languages.  The novelty of our approach is in handling of
  liveness  of closures,  an  issue  that arises  due  to laziness  at
  runtime.

@@

A notion  that generalizes liveness  is {\em demand}.   While liveness
gives the patterns of future uses of a variable, demand represents the
future uses of the value of an expression.  The demand on an expression
$e$   is   also   a   set    of   access   paths---the   subset   of
$\{\acar,\acdr\}^\ast$ which the  context of $e$ may  explore of $e$'s
result.   To see  the need  for demands,  consider the expression
$(\LET~x\leftarrow  (\CDR~y)~\IN~   (\SRETURN~x))$.   Assume   that  the
context of this expression places  the demand $\lbrace \epsilon, \acar
\rbrace$. Since the  value of the expression is the  value of $x$, the
demand translates to the liveness  $[x \mapsto \lbrace \epsilon, \acar
  \rbrace]$.   Due  to  the  \LET\   definition  which  binds  $x$  to
$(\CDR~y)$, the liveness of $x$  now becomes the demand on $(\CDR~y)$.
This, in  turn, generates  the liveness $\lbrace  y.\epsilon, y.\acdr,
y.\acdr\acar  \rbrace$.   These are  the $y$-rooted  accesses
required to  explore $\lbrace  \epsilon, \acar  \rbrace$ paths  of the
result of $(\CDR~y)$.
%% As an  analogy with classical  (strong) liveness analysis,  $y$ and
%% $z$ are live at the entry $\pi: x:=y+z$, if and only if $x$ is live
%% at  exit  of $\pi$.   In  our  terminology, the  liveness  $\lbrace
%% \epsilon\rbrace $ of $x$ at the  exit from $\pi$ becomes the demand
%% on  $y+z$,  and  this,  in turn  generates  the  liveness  $\lbrace
%% y.\epsilon, z.\epsilon \rbrace$ at the entry of $\pi$.

@@

\begin{figure*}[t!]
\begin{center}\footnotesize
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{|@{}c@{\ }|@{\ }c@{}|@{\ }c@{\ }|}
\hline
Premise & Transition & Rule name \\ 
\hline
\hline 
\makecell[t]{ $\rho(x) \mbox{ is } \bot$} & $\rho, S,
  \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow \bang$ &
{\sc car-bang} 
\\
\hline
\makecell[t]{$\heap(\rho(x)) \mbox{ is } (\langle s, \rho'\rangle, d)$} & \makecell[t]{$\rho,\, S,\,  \heap,\,
(\CAR~x), \cred{\sigma} \rightsquigarrow$ $ \rho', \,(\rho, addr(\langle
s, \rho'\rangle), (\CAR~x),\cred{\sigma} )\!:\!S, \heap,\, s, \, \cred{\sigma}$ }     &
{\sc car-1-clo} 
\\
\hline
\makecell[t]{ $\heap(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$} & 
\makecell[t]{ $\rho, S, \heap, (\CAR~x), \cred{\sigma}
  \rightsquigarrow$ 
  $\rho', (\rho, \rho(x), (\CAR~x), \cred{\sigma})\!:\!S, \heap, s,\cred{\renewcommand{\arraystretch}{1}\left\{
    \begin{array}{lr}
      {\emptyset}&\mbox{if } \sigma = \emptyset \\       
      {\{\epsilon\} \cup \acar\sigma }&\mbox{otherwise}
    \end{array}\right.
  }$   }      
&
{\sc car-clo}\\
\hline
\makecell[t]{$\cred{GC(\rho_1, S_1, \heap_1, (\LET~x\leftarrow
    s~\IN~e), \sigma) = (\rho, S, \heap)}$,\\$\ell$ is a new location}& \makecell[t]{$\rho, S, \heap, (\LET~x\leftarrow
  s~\IN~e), \cred{\sigma}$  $ \rightsquigarrow \rho\oplus[x
    \mapsto \ell], S, \heap[\ell := \langle s, \lfloor\rho\rfloor_{FV(s)}, \sigma_x\rangle], e, \cred{\sigma}$ \\
    where $\sigma_x\ =\  \lfloor\mathcal{L}(e,\sigma,\Lfonly)\rfloor_{\{x\}}$} &
{\sc let} \\ 
\hline
\end{tabular}
\end{center}
\caption{Minefield  semantics for  \CAR\   and \LET.
  \warning{Add a note mentioning that we handle the issue of requiring
    a      specific     demand      by      introducing     a      new
    symbol}\label{fig:minefield-semantics-for-some}}
\vspace*{-2mm}
\end{figure*}

@@

\begin{figure}[t]
  \begin{minipage}{.45\columnwidth}
    \begin{align*}
      \nont{i}{\acar} &\rightarrow \acar\nont{j}{\acar}\bcar\nont{\mathit{all}}{} \\
      \nont{i}{\acar} &\rightarrow \clazy\nont{j}{\acar}\bcar\nont{\mathit{all}}{} \\
      \nont{\mathit{all}}{} &\rightarrow \epsilon \mid \acar\nont{\mathit{all}}{} \mid
      \acdr\nont{\mathit{all}}{}
    \end{align*}
  \end{minipage}
  \begin{minipage}{.45\columnwidth}
    \input{equivalent-program}
  \end{minipage}\\
  \mbox{}\hfill(a)\hfill(b)\hfill\mbox{}
  \caption{A possible grammar generated by the proof of
    Lemma~\ref{lemma:grammar-undecidable} and a program to realize the
  grammar.}\label{fig:turing-pgm}
\vspace*{-6mm}
\end{figure}

@@

\begin{figure*}[t!]
\renewcommand{\arraystretch}{.1}
\begin{tabular}{@{}c@{}@{}c@{}@{}c@{}}
   \hskip -4mm{\epsfig{file=fibheap.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=sudoku.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=nperm.eps, height=\hgt}}
\\ \hskip -4mm{\epsfig{file=paraffins.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=lcss.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=huffman.eps, height=\hgt}}
\\ \hskip -4mm{\epsfig{file=knightstour.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=nqueens.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=deriv.eps, height=\hgt}}
\\ \hskip -4mm{\epsfig{file=treejoin.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=lambda.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=gc_bench.eps, height=\hgt}}
%\\ (a) & (b)
\end{tabular}
\vspace*{-8mm}
 \caption{Memory usage.  
The \RGCLine\ and the \LGCLine\ curves indicate the number of cons
cells  in  the  active   semi-space  for  RGC  and  LGC
respectively.  The \ReachLine\ curve represents the number of
reachable cells and the  \UseLine\ curve represents the
number  of  cells  that  are actually  live  (of  which
liveness analysis does a static approximation).  x-axis
is the time measured  in number of cons-cells allocated
(scaled down by factor $10^5$). y-axis is the number of
cons-cells (scaled down by $10^3$).}
\label{fig:memory-usage} \figrule
\vspace*{-1mm}
\end{figure*}

\vspace*{-6.5mm}
\section{The GC scheme}
\label{sec:GC-scheme}
